---
title: Murder tables!?
author: Stat 251
categories:
- slides
- Week11
engine: knitr
format:
  revealjs:
    navigation-mode: vertical
    logo: N.svg
    theme: inverse.scss
    includes:
      in_header: header.html
date: '2026-04-03'
---

```{r setup}
library(tidyverse)
```

## Logistics 

- Submit homework 8 by midnight! Speak now with further questions or ...

- Focus  on homework 9: Murder in SQL City


## Homework 9 


- Solve a murder!


```{=html}
<iframe width="100%" height = "500px" src="https://unl-statistics.github.io/stat251/homework/09-join.html"></iframe>
```

## Connection of homework 9 to joins

```{r db-setup-r}
library(RSQLite)
library(DBI)

# Your database connection code goes here. Store the connection as con 
con <- dbConnect(RSQLite::SQLite(), "../homework-repos/09-join/sql-murder-mystery.db")

table_names <- c("crime_scene_report", 
                 "drivers_license", 
                 "facebook_event_checkin",
                 "get_fit_now_check_in", 
                 "get_fit_now_member", 
                 "income", 
                 "interview",
                 "person")

# For each table name, read the tables in and store them as the name of the table
purrr::walk(table_names, function(x) {
  assign(x, dbReadTable(con, x), envir = .GlobalEnv)
})

# close the connection when you're done
dbDisconnect(con)
```
- Make sure to look at the relationship between the different data sets!

- The first filter statement gives you enough information to start searching for witness statements. 

```{r}
crime_scene_report %>% filter(city=="SQL City", date == 20180115, type=="murder") %>% pluck("description") %>% cat()
```
- Eventually you will be in the situation that you need to look up information from a different table based on a key variable. Instead of using a filter with that information, merge the two tables to directly incorporate the information from the other table.
